---
title: "Script for filtering transcription factors in MaxQuant proteingroups table"
output: github_document
---

Load necessary libraries.
```{r warning=FALSE}
library(data.table)
library(xlsx)
library(biomaRt)
```

Select rat BioMart database for later use.
```{r}
mart_rat <- useMart(biomart = "ensembl", dataset = "rnorvegicus_gene_ensembl")
```

Generate list of transcription factors based on "DNA-binding transcription factor activity (GO:0003700)" and "RNA polymerase II cis-regulatory region sequence-specific DNA binding (GO:0000978)" categories with mammalian filter from www.geneontology.org (data downloaded 16.03.2020).
```{r}
#Read in the Gene Ontology files.
TFs <- fread("20200316_GENEONTOLOGY_TRANSCRIPTION_FACTOR_MAMMALIAN.txt", header = FALSE)
RNAPII <- fread("20200316_GENEONTOLOGY_RNAPOLII_SEQUENCE-SPECIFIC_DNA_BINDING_MAMMALIAN.txt", header = FALSE)

#Convert gene symbols uppercase and keep only unique records.
combined_TFs <- unique(c(toupper(TFs$V2), toupper(RNAPII$V2)))
```

Main function for filtering rows containing transcription factors.
```{r}
Filter_TFs <- function(input, sheetname) {
  
#Read the indicated sheet from the input xlsx file and convert it to data.table format.
dt <- as.data.table(read.xlsx(input,sheetName = sheetname))

#Split Protein IDs groups, making new row for each individual Protein ID.
dt_protein_IDs <- dt[,list(UniprotIDs_split = unlist(strsplit(as.character(Protein.IDs), ";"))), by=eval(colnames(dt))]
#Convert Uniprot IDs to gene symbols using BioMart.
UniprotIDs <- dt_protein_IDs$UniprotIDs_split
ID_to_name <- getBM(attributes = c('external_gene_name', "uniprot_gn_id"), filters = "uniprot_gn_id", values = UniprotIDs, mart=mart_rat)
dt_protein_IDs <- merge(dt_protein_IDs, ID_to_name, by.x ="UniprotIDs_split", by.y = "uniprot_gn_id")
#Find the values of Protein IDs column in rows where the converted gene name is contained in the transcription factor list.
split_IDs_TF <- dt_protein_IDs[toupper(external_gene_name) %in% combined_TFs]$Protein.IDs

#Split Gene names groups, making new row for each individual Gene name.
dt_gene_symbol <- dt[,list(Genes_split = unlist(strsplit(as.character(Gene.names), ";"))), by=eval(colnames(dt))]
#Find the values of Gene names column in rows where the split gene name is contained in the transcription factor list.
split_names_TF <- dt_gene_symbol[toupper(Genes_split) %in% combined_TFs]$Gene.names

#Based on the lists in split_IDs_TF and split_names_TF, keep only the rows from the original table that contain at least one transcription factor.
TF_list <- dt[Protein.IDs %in% split_IDs_TF | Gene.names %in% split_names_TF]

return(TF_list)
}

```

Apply the filtering function on the MaxQuant proteingroups tables.
```{r}
Filtered_replicate1 <- Filter_TFs("Supplementary File 3.xlsx", "Replicate1")
Filtered_replicate2 <- Filter_TFs("Supplementary File 3.xlsx", "Replicate2")

#Write the filtered tables to files.
fwrite(Filtered_replicate1, "TFs_replicate1.csv")
fwrite(Filtered_replicate2, "TFs_replicate2.csv")
```

Information about the R session.
```{r}
sessionInfo()
```

